A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-09 latex/main.tex
WIP on RIPPLES

\subsection*{Comparison with RIPPLES}

RIPPLES~\cite{Turakhia2022} is a method for detecting 
recombination events using a tree inferred by UShER.
Recombination is detected by first identifying plausible branches
carrying multiple mutations (at least 3, by default).
For each candidate branch,
RIPPLES then systematically splits the sequence into 2 or 3 segments, and
finds parsimonious placements for each segment independently.
A recombination event at that node is then flagged if
the overall parsimony of these multiple placements
sufficiently improves over the original single-parent origin in the tree
(the ``parsimony improvement'' parameter, $p=3$, by default).
RIVET~\cite{Smith2023tracking} is a tool for quality control
and visualization of these recombination events.

@@

@@

>>>>

\subsection*{Comparison with RIPPLES}
RIPPLES~\cite{Turakhia2022} is a method for detecting 
recombination events using a tree inferred by UShER,
and is the only method directly comparable with sc2ts.
Recombination is detected by first identifying plausible branches
carrying multiple mutations .
For each candidate branch,
RIPPLES then systematically splits the sequence into 2 or 3 segments, and
finds parsimonious placements for each segment independently.
A recombination event at that node is then flagged if
the overall parsimony of these multiple placements
sufficiently improves over the original single-parent origin in the tree
(the ``parsimony improvement'' parameter, $p=3$, by default).
RIVET~\cite{Smith2023tracking} is a tool for quality control
and visualization of these recombination events.

We ran fast-RIPPLES\cite{Smith2023tracking} on the Viridian UShER tree
subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
At $p=3$, this resulted in a total of 4111 recombination events
of which 595 could be mapped to one of the 855 recombination events in the 
sc2ts ARG.

Figure \ref{fig:ripples_sc2ts_events}A). Analysis of the descendant 
samples showed that many Pango X lineages were not associated with
recombination events (including XBB), and we therefore ran
RIPPLES with the more lenient value of $p=3$ (the default).
This resulted in a total of 4111 recombination events
of which 595 have corresponding recombination events in the 
sc2ts ARG (Figure \ref{fig:ripples_sc2ts_events}B).
Analysis of the descending samples for each event shows that 
10 Pango of the 44 X lineages present in the dataset
% NOTE: listing them like this for sorting and manipulation purposes
(%
XA, 
XAC, 
XAE, 
XAJ,
XB, 
XBB, 
XBM, 
XC, 
XL, 
XW,%
)
are associated unequivocally with a single recombination event.

XBR, XBD 
XS appears twice

Another event suggests that 
XAF 
XAL, 
XE, 
XH, 
XJ, 
XM, 
XY, 
and 
82 BA.2 samples share a recombination event [With what breakpoint and 
parents?].

@@

Figure \ref{fig:ripples_sc2ts_events}A). 
See Document\ref{sec:detailed_ripples} and Table~\ref{tab:ripples_x_lineages}
for detailed analysis.

@@

\subsection{Detailed RIPPLES analysis}
\label{sec:detailed_ripples}
